Assignment 2: Interpreting Quantitative Findings

### 2022-12-12 paper.tex
final edits

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Assignment 2: Interpreting Quantitative Findings},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{lastpage}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead{}
\renewcommand{\headrulewidth}{0pt}
\fancyfoot[CO,CE]{Page \thepage\ of \pageref{LastPage}}
\usepackage{floatrow}
\floatsetup[figure]{capposition=top}
\floatsetup[table]{capposition=top}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Assignment 2: Interpreting Quantitative Findings}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{\hfill\break
University of Glasgow\\
\strut \\
Student ID: 2819052\\
Course: Quantitative Methods\\
Number of words: 2481}
\author{}
\date{\vspace{-2.5em}}

@@

In 1998, the Good Friday/Belfast Agreement was signed and later
validated by the voters, and this agreement marked a new era of hope
towards peace, equality, and inclusion in Northern Ireland
(\protect\hyperlink{ref-galligan2013gender}{Galligan 2013}). The main
focus on the agreement and referendum was obviously to stop the violent
conflict and seek peace, but one part of the agreement also emphasized
an equality agenda (\protect\hyperlink{ref-Hayward2021}{Hayward 2021}).
Thus, the agreement also marked the first Northern Irish formal
recognition of women's rights to political inclusion
(\protect\hyperlink{ref-galligan2013gender}{Galligan 2013}). However, a
formal recognition does not necessarily imply that gender equality
trickles down into social norms and practices. Therefore, this report
examines the contemporary state of women's equality in Northern Ireland.

@@

The causal relationship that we examine is thus a negative relationship.
Although, we call this relationship causal, it is important to clarify
that I do not imply that there is anything biological or deterministic
that reduce women's income in general. Rather, the causal link is
interpreted as a result of the societal discriminatory practices
explained in the previous paragraph. To make a convincing inquiry of
this causal relationship, we must also control for a number of other
factor related to income and gender. This is important to make sure that
our analysis do not simply show a spurious correlation as it could in
fact be another factor that influences both gender and income. These
control variables are introduced in
\protect\hyperlink{data-and-method}{Data and Method}.

@@

Our sample for the analysis consists of data from Northern Ireland Life
and Times Survey (NILT). The data have been collected every year since
1998, but this analysis uses the survey from 2012
(\protect\hyperlink{ref-ark_2012}{ARK 2012}). The respondents for the
NILT survey was chosen from a systematic random sample of addresses.
From this sample of 2126 addresses, 1204 questionnaires were fully
completed using partly face-to-face- and self-completion questionnaires
(\protect\hyperlink{ref-ark_2012}{ARK 2012}). Our sample of analysis is
further reduced as not all respondents have answered our variables of
interest, and we are therefore not able to include those respondents in
our analysis. In the tables below, we present descriptive statistics for
our variables of interest in both the full NILT sample and our sample of
analysis. The first tables shows the categorical variables, where a
frequency distribution are shown
(\protect\hyperlink{ref-fogarty2018quantitative}{Fogarty 2018, 88}). The
second table shows the numerical variables, where we have calculated the
mean and standard deviation
(\protect\hyperlink{ref-fogarty2018quantitative}{Fogarty 2018, 95}). For
a full visualization of all variables included in the analysis, see
\protect\hyperlink{appendix}{Appendix}.

@@

The dependent variable or outcome that we want to explain in this
analysis is income. Although income in itself is not complex concept
compared to other theoretical concepts in social science, it can still
be measured in many different ways. In this analysis, we measure income
by the question: \emph{What is your personal income before tax and
national insurance contributions? (£ per year)}. It is thus a numeric,
most specifically a ratio scale with an value of absolute zero -
although it is quite uncommon to have a personal income of zero
(\protect\hyperlink{ref-fogarty2018quantitative}{Fogarty 2018, 56}).

The independent variable or explanatory variable of our analysis is
gender. Gender is a much more complex concept, and there is a large
discussion about the distinction between gender and sex, and the
biological and cultural components of these
(\protect\hyperlink{ref-unger1993sex}{Unger and Crawford 1993}).
Quantitative research have mostly been dominated by gender being
measured as a binary division between male and female. Thus, we are also
limited to such a measurement of gender in the NILT 2012 survey. In this
analysis, we therefore measure gender by a nominal, binary variable of
either ``male'' or ``female''.

@@

There are several insights from this visualization. The vertical
differences between the descriptive statistics, mean and median
signifies that there is a general income difference between males and
females, and that females in average have lower incomes than men.
Initially, this suggests an acceptance of our alternative research
hypothesis, but we will wait will any conclusion till the analysis,
where we also control for other related factors. Another interesting
insight is that there the distribution of income seems to slightly
skewed. For both males and females, there are some respondents with a
very high income compared to most other respondents. This is also seen
as there is a difference between the mean and the median for both
groups. These respondents have such extreme high incomes that they might
be outliers compared to the general trend, and we need to be aware of
that in our analysis
(\protect\hyperlink{ref-fogarty2018quantitative}{Fogarty 2018,
221--22}).

@@

This report examines the question of whether gender affect income in
contemporary Northern Ireland. As described in the
\protect\hyperlink{introduction}{Introduction}, the analysis therefore
seeks to accept wither the null hypothesis or the alternative hypothesis
- that being a woman rather than a man affects a lower income. As
described in the \protect\hyperlink{data-and-method}{Data and Method}
section, our analysis consists of survey data, where our main variables
are answered gender and personal income. To examine our hypothesis, the
following table shows the results of two regressions. First, our model
is shown as a bivariate regression excluding our control variables, and
secondly our full model is shown including our control variables.

@@

The primary finding from the regression results is that the variable
gender have an significant effect on income. We see that the p-value is
less than 0.01. This indicates that there is indeed a significant
difference between the income of men and women. However, a significant
p-value does not by itself means that the difference is important or
meaningful - we also need to interpret the effect size
(\protect\hyperlink{ref-field2012discovering}{Field, Miles, and Field
2012, 57}). If we take a look at the effect size (\(\beta_1\)), it has a
value of approx -5,068. Thus according to the model, women in average
earns 5,068 pounds a year less than men. That seems quite substantial as
the average income in our sample is 16,892 pounds a year (See Table 2).
Based on our analysis, we therefore accept our alternative hypothesis -
that being a women rather than a man has a significant negative effect
in your income.

In model, where we include our control variables, we see that there is
trade union membership and being a supervisor has a significant positive
effect on income, whereas age and to some extent also having a
homosexual orientation rather than a heterosexual orientation have a
significant negative effect on income. However, the effect of gender is
independent of religion, sexual orientation, constitutional view, trade
union membership, being a supervisor, and age, as including these
control variables into the regression does not render the effect of
gender on income insignificant. If we compare the model with and without
control variables, we see that both R\^{}2 and adjusted R\^{}2 is
increased with indicates that including the control variables does
indeed yield a better fit of data.

Before concluding on our empirical results, it is important to consider
how reliable our regression model is. One relevant concern of our
analysis is the influence of outliers, i.e.~very influential data points
(\protect\hyperlink{ref-fogarty2018quantitative}{Fogarty 2018, 221--22};
\protect\hyperlink{ref-field2012discovering}{Field, Miles, and Field
2012, 190}). As identified in the {[}Data and Methods{]} section, the
dependent variables seems to have a data points with incomes that are
very much higher. Most respondents have an income around 17,000 pounds a
year, while a few respondents have a yearly income around more than
60,000 pounds. It is beyond the scope of this report to examine whether
these data points are in fact significant outliers, but it is important
to notice when interpreting the results.

Another concern to our model is heteroscedasticity, i.e.~whether there
is not a constant variance of residuals on the levels of the dependent
variable, income (\protect\hyperlink{ref-field2012discovering}{Field,
Miles, and Field 2012, 272}). In the plot below, we visualize the fitted
values with the residuals to see if the variances is constant along the
x-axis. Below the plot is also shown the results of a Breusch-Pagan test
to more formally test to heteroscedasticity
(\protect\hyperlink{ref-fogarty2018quantitative}{Fogarty 2018, 228}).

@@

\caption{Scatterplot of Fitted Values and Residuals}\label{fig:unnamed-chunk-4}
\end{figure}

Unfortunately, the plot shows that the variance seem to increase for
higher values of our dependent variables, income, and the Breusch-Pagan
test also suggests that there is heteroscedasticity as the p-value is
lower than 0.001. One solution to this problem could be to re-estimate
our regression model with robust standard errors, but that it has not
been done for this report
(\protect\hyperlink{ref-fogarty2018quantitative}{Fogarty 2018, 229}).

@@

Using a survey-based research design, we employed a multivariate linear
regression analysis to answer this question. Our analysis showed that
there is indeed a gender pay gap in contemporary Northern Ireland. We
found that being a woman compared to a man resulted in average of an
approx. 5000 pounds lower yearly income - independent of factors such as
trade union membership, being a supervisor, age and more. Thus, our
analysis supports the arguments that gender affects income rooted in
discriminatory practices as theory suggests.

However, there are also limitations to our study that it is important to
take into concern. Firstly, there might be a few influential data points
- so-called outliers - that drives the effect to be larger than the
general trend. Secondly, we identified heteroscedasticity in our model,
and especially large incomes seems to have larger residuals. These are
primarily limitations of the reliability of the study, since it concerns
the model and methodological choices
(\protect\hyperlink{ref-bryman2016social}{Bryman 2016, 41}). If these
results can be reproduced with a model with robust standard errors or
after a check for outliers, it would increase the reliability of our
findings. Another points of discussion is the validity of our findings,
and here there is mostly the question of external validity as our sample
is reduced significantly during the data-cleaning process
(\protect\hyperlink{ref-bryman2016social}{Bryman 2016, 41--42}). A
further improvement here could have been to directly analyze if the
respondents in our data are representative compared to the general
population of Northern Ireland.

@@

\caption{Bar plot of Gender}\label{fig:unnamed-chunk-5}
\end{figure}

@@

\caption{Bar plot of Religion}\label{fig:unnamed-chunk-6}
\end{figure}

@@

\caption{Bar plot of Sexual Orientation}\label{fig:unnamed-chunk-7}
\end{figure}

@@

\caption{Bar plot of Constitutional View}\label{fig:unnamed-chunk-8}
\end{figure}

@@

\caption{Bar plot of Trade union membership}\label{fig:unnamed-chunk-9}
\end{figure}

@@

\caption{Bar plot of Supervisor}\label{fig:unnamed-chunk-10}
\end{figure}

@@

\caption{Density plot of Age}\label{fig:unnamed-chunk-11}
\end{figure}

@@

\caption{Density plot of Income}\label{fig:unnamed-chunk-12}
\end{figure}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Assignment 2: Interpreting Quantitative Findings},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{lastpage}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead{}
\renewcommand{\headrulewidth}{0pt}
\fancyfoot[CO,CE]{Page \thepage\ of \pageref{LastPage}}
\usepackage{floatrow}
\floatsetup[figure]{capposition=top}
\floatsetup[table]{capposition=top}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Assignment 2: Interpreting Quantitative Findings}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{\hfill\break
University of Glasgow\\
\strut \\
Student ID: 2819052\\
Course: Quantitative Methods in the Social Sciences\\
Number of words: 2747\\}
\author{}
\date{\vspace{-2.5em}}

@@

\doublespacing

In 1998, the Good Friday/Belfast Agreement was signed and later
validated by the voters, and this agreement marked a new era of hope
towards peace, equality, and inclusion in Northern Ireland
(\protect\hyperlink{ref-galligan2013gender}{Galligan 2013}). The main
focus on the agreement and referendum was obviously to stop the violent
conflict and seek peace, but a part of the agreement also emphasized an
equality agenda (\protect\hyperlink{ref-Hayward2021}{Hayward 2021}).
Thus, the agreement also marked the first Northern Irish formal
recognition of women's rights to political inclusion
(\protect\hyperlink{ref-galligan2013gender}{Galligan 2013}). However, a
formal recognition does not necessarily imply that gender equality
trickles down into societal norms and practices. Therefore, this report
examines the contemporary state of women's equality in Northern Ireland.

@@

The causal relationship that we examine is thus a negative relationship.
Although, we call this relationship causal, it is important to clarify
that I do not imply that there is anything biological or deterministic
that reduce women's income in general. Rather, the causal link is
interpreted as a result of the societal discriminatory practices
explained in the previous paragraph. To make a convincing inquiry of
this causal relationship, we must also control for a number of other
factors related to income and gender. This is important to make sure
that our analysis do not simply show a spurious correlation as it could
in fact be another factor that influences both gender and income. These
control variables are introduced in
\protect\hyperlink{data-and-method}{Data and Method}.

@@

Our sample of analysis consists of data from the Northern Ireland Life
and Times Survey (NILT). The data have been collected every year since
1998, but this analysis uses the survey from 2012
(\protect\hyperlink{ref-ark_2012}{ARK 2012}). The respondents for the
NILT survey was chosen from a systematic random sample of addresses.
From this sample of 2126 addresses, 1204 questionnaires were fully
completed using partly face-to-face- and self-completion questionnaires
(\protect\hyperlink{ref-ark_2012}{ARK 2012}). Our sample of analysis is
further reduced as not all respondents have answered our variables of
interest, and we are therefore not able to include those respondents in
our analysis. In the tables below, we present descriptive statistics for
our variables of interest in both the full NILT sample and our sample of
analysis. The first tables shows the categorical variables, where a
frequency distribution are shown
(\protect\hyperlink{ref-fogarty2018quantitative}{Fogarty 2018, 88}). The
second table shows the numerical variables, where we have calculated the
mean and standard deviation
(\protect\hyperlink{ref-fogarty2018quantitative}{Fogarty 2018, 95}). For
a full visualization of all variables included in the analysis, see
\protect\hyperlink{appendix}{Appendix}.

@@

The dependent variable or outcome that we want to explain in this
analysis is income. Although income in itself is more concrete concept
compared to other theoretical concepts in social science (such as
e.g.~``solidarity''), it can still be measured in many different ways.
In this analysis, we measure income by the question: \emph{What is your
personal income before tax and national insurance contributions? (£ per
year)}. It is thus a numeric, more specifically a ratio scale with a
value of absolute zero - although it is quite uncommon to have a
personal income of zero
(\protect\hyperlink{ref-fogarty2018quantitative}{Fogarty 2018, 56}).

The independent variable or explanatory variable of our analysis is
gender. Gender is a much more complex concept, and there is a large
discussion about the distinction between gender and sex, and the
biological and cultural components of these
(\protect\hyperlink{ref-unger1993sex}{Unger and Crawford 1993}).
Quantitative research have mostly been dominated by gender being
measured as a binary division between male and female. Likewise, we are
also limited to such a measurement of gender in the NILT 2012 survey. In
this analysis, we therefore measure gender by a nominal, binary variable
of either ``male'' or ``female''.

@@

There are several insights from this visualization. The vertical
differences between the descriptive statistics, mean and median
signifies that there is a general income difference between males and
females, and that females in average have lower incomes than men.
Initially, this suggests an acceptance of our alternative research
hypothesis, but we will wait with any conclusion till the analysis,
where we also control for other related factors. Another interesting
insight is that the distribution of income seems to be slightly skewed.
For both males and females, there are some respondents with a very high
income compared to most other respondents. This can also noticed as
there is a difference between the mean and the median for both groups.
These respondents have such extreme high incomes that they might be
outliers compared to the general trend, and we need to be aware of that
in our analysis (\protect\hyperlink{ref-fogarty2018quantitative}{Fogarty
2018, 221--22}).

@@

This report examines the question of whether gender affect income in
contemporary Northern Ireland. As described in the
\protect\hyperlink{introduction}{Introduction}, the analysis therefore
seeks to accept whether the null hypothesis or the alternative
hypothesis - that being a woman rather than a man affects a lower
income. As described in the \protect\hyperlink{data-and-method}{Data and
Method} section, our analysis consists of survey data, where our main
variables are gender and personal income. To examine our hypothesis, the
following table shows the results of two regressions. First, our model
is shown as a bivariate regression excluding our control variables, and
secondly our full model is shown including our control variables.

@@

The primary finding from the regression results is that the variable
gender have an significant effect on income. We see that the p-value is
less than 0.01. This indicates that there is indeed a significant
difference between the income of men and women. However, a significant
p-value does not by itself means that the difference is important or
meaningful - we also need to interpret the effect size
(\protect\hyperlink{ref-field2012discovering}{Field, Miles, and Field
2012, 57}). If we take a look at the effect size (\(\beta_1\)), it has a
value of approx -5,068. Thus according to the model, women in average
earns 5,068 pounds a year less than men. That seems quite substantial as
the average income in our sample is 16,892 pounds a year (See Table 2).
Based on our analysis, we therefore accept our alternative hypothesis -
that being a woman rather than a man has a significant negative effect
in your income.

In the model, where we include our control variables, we see that trade
union membership and being a supervisor has a significant positive
effect on income, whereas age and to some extent also having a
homosexual orientation rather than a heterosexual orientation have a
significant negative effect on income. However, the effect of gender is
independent of religion, sexual orientation, constitutional view, trade
union membership, being a supervisor, and age, as including these
control variables into the regression does not render the effect of
gender on income insignificant. If we compare the model with and without
control variables, we see that both \(R^2\) and adjusted \(R^2\) is
increased with indicates that including the control variables does
indeed yield a better fit of data.

In this way, our empirical findings support the argument that there is
still a gender pay gap in Northern Ireland. Our finding that the gap
between men and women is independent of being a supervisor or not is
particularly interesting. It suggests that the income gap between men
and women exists for both, and the gender pay gap cannot only be
explained by men more often being in leadership positions. In the
following visualization, this finding can be seen clearly.

@@

\caption{Scatterplot of Income, Gender, and Supervisor}\label{fig:unnamed-chunk-4}
\end{figure}

In the visualization, there is also shown two linear regression of
gender on income, and we see that both slopes indicate a negative
correlation - although the intercepts differ.

Before concluding on our empirical results, it is also important to
consider how reliable our regression model is. One relevant concern of
our analysis is the influence of outliers, i.e.~very influential data
points (\protect\hyperlink{ref-fogarty2018quantitative}{Fogarty 2018,
221--22}; \protect\hyperlink{ref-field2012discovering}{Field, Miles, and
Field 2012, 190}). As identified in the {[}Data and Methods{]} section,
the dependent variables seems to have a data points with incomes that
are very much higher. Most respondents have an income around 17,000
pounds a year, while a few respondents have a yearly income around more
than 60,000 pounds. It is beyond the scope of this report to examine
whether these data points are in fact significant outliers, but it is
important to notice when interpreting the results.

Another concern to our model is heteroscedasticity, i.e.~whether there
is not a constant variance of residuals on the levels of the dependent
variable, income (\protect\hyperlink{ref-field2012discovering}{Field,
Miles, and Field 2012, 272}). In the plot below, we visualize the fitted
values with the residuals to see if the variances is constant along the
x-axis. Below the plot is also shown the results of a Breusch-Pagan test
to more formally test for heteroscedasticity
(\protect\hyperlink{ref-fogarty2018quantitative}{Fogarty 2018, 228}).

\caption{Scatterplot of Fitted Values and Residuals}\label{fig:unnamed-chunk-5}
\end{figure}

Unfortunately, the plot shows that the variance seem to increase for
higher values of our dependent variables, income, and the Breusch-Pagan
test also suggests that there is heteroscedasticity as the p-value is
lower than 0.001. One solution to this problem could be to re-estimate
our regression model with robust standard errors, but this has not been
possible within the scope of this report
(\protect\hyperlink{ref-fogarty2018quantitative}{Fogarty 2018, 229}).

@@

Using a survey-based research design, we employed a multivariate linear
regression analysis to answer this question. Our analysis showed that
there is indeed a gender pay gap in contemporary Northern Ireland. We
found that being a woman compared to a man resulted in average of an
approx. 5000 pounds lower yearly income - independent of factors such as
trade union membership, being a supervisor, age and more. Here, it is
particularly interesting that the effect of gender on income is
independent of having supervision, because it suggests that the gender
inequality exists across having a leadership position or not. Thus, our
analysis supports the arguments that gender affects income rooted in
discriminatory practices as theory suggests.

However, there are also limitations to our study that it is important to
take into concern. Firstly, there might be a few influential data points
- so-called outliers - that drives the effect to be larger than the
general trend. Secondly, we identified heteroscedasticity in our model,
and especially large incomes seems to have larger residuals. These are
primarily limitations of the reliability of the study, since it concerns
the model and methodological choices
(\protect\hyperlink{ref-bryman2016social}{Bryman 2016, 41}). If these
results can be reproduced with a model with robust standard errors or
after a check for outliers, it would increase the reliability of our
findings. Another point of discussion is the validity of our findings,
and here there is mostly the question of external validity as our sample
is reduced significantly during the data-cleaning process
(\protect\hyperlink{ref-bryman2016social}{Bryman 2016, 41--42}). A
further improvement here could have been to directly analyze if the
respondents in our data are representative compared to the general
population of Northern Ireland.

@@

@@

\caption{Bar plot of Gender}\label{fig:unnamed-chunk-6}
\end{figure}

@@

\caption{Bar plot of Religion}\label{fig:unnamed-chunk-7}
\end{figure}

@@

\caption{Bar plot of Sexual Orientation}\label{fig:unnamed-chunk-8}
\end{figure}

@@

\caption{Bar plot of Constitutional View}\label{fig:unnamed-chunk-9}
\end{figure}

@@

\caption{Bar plot of Trade union membership}\label{fig:unnamed-chunk-10}
\end{figure}

@@

\caption{Bar plot of Supervisor}\label{fig:unnamed-chunk-11}
\end{figure}

@@

\caption{Density plot of Age}\label{fig:unnamed-chunk-12}
\end{figure}

{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/unnamed-chunk-13-1}

\caption{Density plot of Income}\label{fig:unnamed-chunk-13}
\end{figure}
